---
title: "Air Traffic Delays Analysis"
author: "Sumanth"
date: "4/13/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The data has been extracted from the  Marketing Carrier On-Time Performance (Beginning January 2018) data table of the "On-Time" database from the TranStats data library. (Ref: https://www.transtats.bts.gov/)

**I have performed analysis on California State**

Importing libraries
```{r, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
```

Load the data :
```{r,message=FALSE,warning=FALSE}
California_delays <- read_csv("~/CLASSROOM/Analytics inn practice/ASS 2/Sample_CA_airtraffic_delays.csv")
```

### Exploratory Data Analysis 
```{r}
ArrDep_delays <- California_delays %>% select("DayofMonth","DayOfWeek","Origin","DepDelay","DepDelayMinutes","DepDel15","Dest","ArrDelay","ArrDelayMinutes","ArrDel15","Distance",)
ArrDep_delays$DayofMonth<-as.Date(ArrDep_delays$DayofMonth, origin = "2018-01-01")
ArrDep_delays$DayOfWeek<-as.factor(ArrDep_delays$DayOfWeek)
ArrDep_delays$DepDel15<-as.factor(ArrDep_delays$DepDel15)
ArrDep_delays$ArrDel15<-as.factor(ArrDep_delays$ArrDel15)
```

**1. What is the pattern of arrival traffic and departure traffic delays with respect to days and weeks?**
```{r}
ArrDep_delays[is.na(ArrDep_delays)]<-0
Week <- ArrDep_delays %>% 
        group_by( DayOfWeek ) %>% 
        summarise( Count = n(), Avg_ArrDelay=mean(ArrDelay), Avg_DepDelay=mean(DepDelay))
Month <-ArrDep_delays %>% 
        group_by( DayofMonth ) %>% 
        summarise( Count = n(), Avg_ArrDelay=mean(ArrDelay), Avg_DepDelay=mean(DepDelay))
```


```{r}
 ggplot(Week, aes(x=DayOfWeek, y=Avg_ArrDelay)) + geom_bar(stat="identity", fill="steelblue",width = 0.5)+
  geom_text(aes(label=round(Avg_ArrDelay, 2)), vjust=1.6, color="white",
            position = position_dodge(0.9), size=3.5)+
  scale_fill_brewer(palette="Paired")+
 theme_minimal()+labs(title="Average Arrival Delay per Day of Week")
```

From the above graph, It is clearly evident that the Arrival delay for day 1, 5 are high compared to the rest. This analysis suggests there is a greater scope to improve on those days.
```{r}
ggplot(data=Week, aes(x=DayOfWeek, y=Avg_DepDelay)) +geom_bar(stat="identity", fill="#CC6666", width = 0.5)+
  geom_text(aes(label=round(Avg_DepDelay, 2)), vjust=1.6, color="white",
            position = position_dodge(0.9), size=3.5)+
  scale_fill_brewer(palette="Paired")+
  theme_minimal()+labs(title="Average Departure Delay per Day of Week")
```

The Departure delays in the California state at all the origins is higher compared to the arrival departure delays. 

* There are various factors that influence delays. eg: carrier delay, security delay, weather delay etc.,.
* The departure delay is much higher on days 1, 5 and is equally correlated with the arrival delay

```{r}
ggplot(data=Month, aes(x=DayofMonth, y=Avg_ArrDelay)) +geom_bar(stat="identity", fill="steelblue")+
  geom_text(aes(label=round(Avg_ArrDelay, 1)), vjust=1.6, color="white",
            position = position_dodge(0.9), size=2)+
  scale_fill_brewer(palette="Paired")+
  theme_minimal()+labs(title="Avg Arrival Delay by Day of Month")
```


From the above plot it could be inferred that week 1 has the highest the highest number of delays i.e., from jan 01 to jan 08.
```{r}
ggplot(data=Month, aes(x=DayofMonth, y=Avg_DepDelay)) +geom_bar(stat="identity", fill="#CC6666")+
  geom_text(aes(label=round(Avg_DepDelay, 1)), vjust=1.6, color="white",
            position = position_dodge(0.9), size=2)+
  scale_fill_brewer(palette="Paired")+
  theme_minimal()+labs(title="Average Departure Delay by Day of Month")
```

If we observere the trend, the Departure delays are more frequent and higher in the starting 15 days than the rest. 

**2. Can you interpret the traffic delays?**
```{r}
ArrDep_delays %>%
 filter(!is.na(DepDelay)) %>%
 filter(!is.na(ArrDelay)) %>%
 filter(!is.na(ArrDel15)) %>%
 ggplot() +
 aes(x = ArrDelay, y = DepDelay, colour = ArrDel15) +
 geom_point(size = 0.5) +
 scale_color_hue() +
 labs(x = "Arrival Delay(In Min)", y = "Departure Delay(In Min)", title = "Delay Pattern between Arrival & Departures") +
 theme_minimal()

```

* The pattern of the traffic delays data is normally distributed. 
* In specific, The probability of departure delay is higher when there is an early arrival of flights than scheduleed.

  
**3. Which Airport ('Origin Airport') has highest departure delay?**
```{r}
Highest_ArrDep<-ArrDep_delays %>% 
        group_by( Origin ) %>% 
        summarise( Count = n(), Avg_ArrDelay=mean(ArrDelayMinutes), Avg_DepDelay=mean(DepDelayMinutes))
```

```{r}
ggplot(Highest_ArrDep, aes(x=Origin, y=Avg_DepDelay)) +geom_bar(stat="identity", fill="#CC6666",width = 0.2)+
  theme_minimal()+labs(title="Average Departure Delay by Airport")
```

The SFO (San Francisco International Airport) has highest number of Departure delays among all the airports in the caliifornia State.

I have also created Boxplots to look at the distribution of the data for a better visual understanding.
```{r}
ArrDep_delays %>%
 filter(DepDelay >= 0 & DepDelay <= 403L & !is.na(DepDelay)) %>%
 ggplot() +
 aes(x = Origin, y = DepDelay, fill = Origin) +
 geom_boxplot(width = 0.5) +
 scale_fill_hue() +
 labs(x = "Origin Airport", y = "Departure Delay(in min)", title = "Departure Delays per Airport") +
 theme_minimal() +
 theme(legend.position = "none")

```

**4. Which Airport has highest Arrival delay?**
```{r}
ggplot(Highest_ArrDep, aes(x=Origin, y=Avg_ArrDelay)) +geom_bar(stat="identity", fill="#0c4c8a",width = 0.2)+
  scale_fill_brewer(palette="Paired")+
  theme_minimal()+labs(title="Average Arrival Delay by Airport")
```

The SMF (Sacramento International Airport) has the highest Departure delays in the state followed by SJC(San Jose International Airport)

Below, We can find the distribution the data with boxplots.
```{r}
ArrDep_delays %>%
 filter(ArrDelay >= -64L & ArrDelay <= 405L & !is.na(ArrDelay)) %>%
 ggplot() +
 aes(x = Origin, y = ArrDelay, fill = Origin) +
 geom_boxplot(width = 0.3) +
 scale_fill_hue() +
 labs(x = "Airports", y = "Arrival Delay(in min)", title = "Arrival Delays at Airports") +
 theme_minimal() +
 theme(legend.position = "none")
```

**5. How do you relate the delay pattern to the distance travelled?**
```{r}
ggplot(ArrDep_delays, aes(ArrDelay, Distance)) + 
        geom_point(color='steelblue',size = 0.1)+theme_minimal()+
        facet_wrap(~Origin)+
        labs(title="Distance vs Arrival Delays")
```


```{r}
ggplot(ArrDep_delays, aes(DepDelay, Distance)) + 
        geom_point(color='steelblue',size = 0.1)+theme_minimal()+
        facet_wrap(~Origin)+
        labs(title="Distance vs Departure Delays")
```

There is no relation between the delays to distance.

**6. Is there any correlation between weather delay and carrier delay?** 

According to the data, The correlation is almost close to zero by which we can conclude there is no correlation.
```{r}
traffic_delay<-California_delays %>% select('DayofMonth','CarrierDelay','WeatherDelay','NASDelay','SecurityDelay','LateAircraftDelay')                  

traffic_delay[is.na(traffic_delay)]<-0               

 ggplot(traffic_delay) +
 aes(x = CarrierDelay, y = WeatherDelay) +
 geom_point(size = 0.5, colour = "blue") +
 labs(title = "Carrier Vs Weather Delay") +
 theme_minimal() + annotate("text", x = 300, y = 500, label = "correlation = -0.04256")
```
**7. What is the delay pattern you can find in respective states?**

As mentioned, Due to to limited capability of my PC, I have done my analysis on California state only. Airports with Highest number of flights count tend to have more number of delays when compared to other airports.

**8. How many delayed flights were cancelled? (approximation)**
```{r}
#filtering the data :
cancel_delays<-California_delays %>% select('DepDelay','ArrDelay','Cancelled','Diverted')
cancel_delays %>% filter(Cancelled==1 & !is.na(Cancelled) & DepDelay > 0 & !is.na(DepDelay)) %>% count(Cancelled)
```

There are a total number of 44 flights cancelled due to delay.

**9. How many delayed flights were diverted? (approximation)**
```{r}
cancel_delays %>% filter(Diverted==1 & !is.na(Diverted) & DepDelay > 0 & !is.na(DepDelay)) %>% count(Diverted)
```
A total number of 3 flights were diverted due to delay.

**10. What time of the day do you find Arrival delays?**
```{r}
time_delays<-California_delays %>% select('CRSDepTime','DepDelay','CRSArrTime','ArrDelay')

time_delays %>%
 filter(ArrDelay >= 0 & ArrDelay <= 410L & !is.na(ArrDelay)) %>%
 ggplot() +
 aes(x = CRSArrTime, y = ArrDelay) +
 geom_line(size = 0.5, colour = "#0c4c8a") +
 labs(x = "Time (hh:mm)", y = "Arrival Delay", title = "Times of The Day Vs Arrival Delays") +
 theme_minimal()
```

The Above plot is the time series visual graph wiwth respect to Arrival delays. As we can infer, The Afternoons of a day is usually having Arrival delays i.e., from 12:00 pm to 18:00 pm

**11. What time of the day do you find Departure delays?**
```{r}
time_delays %>%
 filter(DepDelay >= 0 & DepDelay <= 417L & !is.na(DepDelay)) %>%
 ggplot() +
 aes(x = CRSDepTime, y = DepDelay) +
 geom_line(size = 0.5, colour = "#CC6666") +
 labs(x = "Time (hh:mm)", y = "Departure Delay", title = "Times of The Day Vs Departure Delays") +
 theme_minimal()
```
 
 The Afternoons of a day is usually having Arrival delays i.e., from 12:00 pm to 18:00 pm
